---
output: github_document
editor_options: 
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

[![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.8223668.svg)](https://doi.org/10.5281/zenodo.8223668)


# SimPoly

The goal of SimPoly is to simulate a EU-PoMS like datasets with known parameters and appropriate built in stochasticity. 

## Installation

``` {r}
# install.packages("devtools")
devtools::install_github("ibartomeus/SimPoly")
```


## Example

```{r}
library(SimPoly)
set.seed(32468)
```


This is a basic example to show how we build one dataset:

First we define the number of species, sites and years to simulate. We do this using the `define_sites_years()` functions. The `pool` argument is a call to `sp_pool()`, which simply defines the number of species (here `pool = 100`), and the occupancy of the `rarest` and `commonest` species. 
```{r example, eval = FALSE, echo=TRUE, message=FALSE, warning=FALSE}
site_years <- define_sites_years(pool = sp_pool(pool = 100, 
                                                rarest = 0.05, commonest = 0.7),
                                 n_years = 3, n_sites = 10)
```

Let's take a closer look at the data.
```{r}
str(site_years)
head(site_years)
length(unique(site_years$siteID))
length(unique(site_years$species))
```

Note that the number of species in the dataset is only `89`, compared with `100` in the species pool. This reflects stochasticity in how the species are assigned to sites.

Second, we specify species attributes such as phenology, abundance and detectability
```{r example, eval = FALSE, echo=TRUE, message=FALSE, warning=FALSE}
pars <- sp_responses(site_years = site_years,
                     pheno_peak_mean = 120, pheno_peak_sd = 50,
                     pheno_range_mean = 25, pheno_range_sd = 5,
                     trend_max = 0, trend_min = -2)
str(pars)
head(pars)
```

There is one row per species, with the following columns:
`opt` = date of maximum abundance
`tol` = spread #range or sdev?
`h` = species abundance. #Not sure if this is the abundance at every site ...
`slope` = trend in abundance per year. # on the abundance scale, or log abundance?
`detect` = detectability. # per pan trap or per individual?

Third, we sample the true abundance values expected at each sampling point.
```{r example, eval = FALSE, echo=TRUE, message=FALSE, warning=FALSE}
dat <- true_abundance(rounds = 8,
                      site_years = site_years,
                      sp_responses = pars)
str(dat)
head(dat)
```

There are 7414 observations: one per species per site.
`year`: an integer from 1-3 (as defined by n_years, above)
`siteID`: which site is it
`round`: month of the year of this survey
`species`: which species
`abundance`: abundance during that visit # presumably this changes through the season, but I can't see how

From this table we can calculate species richness per site
```{r}
(rich <- dcast(dat, siteID ~ "richness", value.var = 'species', function(x) length(unique(x))))
```


Finally, we sample with detection error from the true values.
```{r example, eval = FALSE, echo=TRUE, message=FALSE, warning=FALSE}
dat_obs <- obs_abundance(true_abundance = dat, sp_responses = pars)
head(dat_obs)
```



```{r example, eval = FALSE, echo=TRUE, message=FALSE, warning=FALSE}

#We can simulate a second transect
dat_obs$obs2 <- obs_abundance(true_abundance = dat, sp_responses = pars)$obs #note order is preserved
head(dat_obs)
plot(dat_obs$obs, dat_obs$obs2) #nice expected correlation.
```

```{r example, eval = FALSE, echo=TRUE, message=FALSE, warning=FALSE}

#Or pan traps. Let's assume a different detectability
pars_pantrap <- pars
pars_pantrap$detect <- runif(n = length(pars_pantrap$detect)) #this assume independent detectabilities
dat_obs$obs_pantrap <- obs_abundance(true_abundance = dat, sp_responses = pars_pantrap)$obs #note order is preserved
plot(dat_obs$obs, dat_obs$obs_pantrap) #nice expected correlation, but more noisy.

#this is the final output
head(dat_obs)
```

```{r example, eval = FALSE, echo=TRUE, message=FALSE, warning=FALSE}

plot(dat_obs$abundance, dat_obs$obs, las = 1, xlim = c(0,max(dat_obs$abundance)), 
     ylim = c(0,max(dat_obs$abundance))) 
#you can summarize observed variables per species, site and year
s_dat <- summary_poms(dat_obs, var_name = "obs")
head(s_dat)
scatter.smooth(s_dat$obs_abund ~ s_dat$year, las = 1)
#This can be easily loop to obtain several simulations
```



## Citation

If using this package, please cite it:
```{r comment=NA}
citation("SimPoly")
```

## Acknowledgements

STING and the coenocliner package.
